Inverse conformational selection in lipid-protein binding

### 2021-07-07 manuscriptPGPE.tex
Updates based on referees' comments.

Structures of protein-bound lipids are available in the Protein Data Bank (PDB, \url{http://www.rcsb.org/}) \cite{berman00},
and crystal structures of lipids have been determined \cite{buldt81,pascher92},
but their relation to lipids in bulk membranes in the liquid lamellar phase remains unclear \cite{marsh13b}. Similarly to disordered proteins~\cite{sormanni17}, lipids sample conformational ensemble which is composed of the set of individual structures whose probability to occur is determined by the Boltzmann distribution. In liquid bilayer state, individual phosphatidylcholine lipid samples the conformational ensemble within nanosecond timescale~\cite{ferreira15}. The most accurate experimental information on conformational ensembles of lipids
in this biologically relevant phase are typically derived from NMR experiments, particularly from the
C--H bond order parameters, $S_\mathrm{CH}=\frac{3}{2} \langle \cos^2 \theta -1 \rangle$, where $\theta$ is the angle between C-H bond and membrane normal, and the average is taken over conformational ensemble of lipids~\cite{seelig77c,davis83,Semchyschyn04}.
Because these order parameters are similar as in living cells \cite{gally81,scherer87,seelig90},
the model membranes can be used to resolve the lipid headgroup conformations in biological conditions.
According to these experiments, the glycerol backbone conformations are largely similar irrespective of the headgroup \cite{gally81}, and
the headgroup conformations are similar in the phosphatidylcholine (PC), phosphatidylethanolamine (PE), and phosphatidylglycerol (PG) lipids,
whereas the phosphatidylserine (PS) headgroup is more rigid \cite{wohlgemuth80,buldt81}. 
Notably, however, the signs of $S_\mathrm{CH}$ are not accessible from $^2$H\,NMR experiments \cite{ollila16},
and universal models to map order parameters to structural ensembles are not available \cite{pezeshkian18,akutsu20}.
Therefore, it is not clear if the lipid headgroups in a liquid lamellar bilayer exchange between a few restricted conformations, or can fluctuate freely across a wide conformational range.

@@

Our results suggest that lipid headgroup binding to proteins can be described by the {\it inverse conformational selection model}
where unbound lipids sample a wide conformational ensemble offering the large selection of conformations
that can bind to various binding sites in substrates, such as protein, drug, RNA, or virus, that remain conformationally fixed.
In this model, the binding energy is dominated by
intermolecular interactions and changes in lipid conformational entropy, while 
conformational restrictions inherent to the lipids play a minor role.
Here we have focused only on lipid headgroups and potential applications of the proposed model on interactions between hydrophobic acyl chains and proteins are left for future studies. 
Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling
and functions of lipid regulated membrane proteins in general. 
In bionanotechnology, our model facilitates, for example,
the design of phoshopolipid-specific antibodies and lipid nanoparticle carriers with minimal sides effects for mRNA vaccines.
Furthermore, our work has demonstrated the power of open access MD simulation data
to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.

>>>>

Structures of protein-bound lipids are available in the Protein Data Bank (PDB, \url{http://www.rcsb.org/}) \cite{berman00},
and crystal structures of lipids have been determined \cite{buldt81,pascher92},
but their relation to lipids in bulk membranes in the liquid lamellar phase remains unclear \cite{marsh13b}. Similarly to disordered proteins~\cite{sormanni17}, lipid molecules sample conformational ensembles which are composed of sets of individual structures whose probabilities of occurrences are determined by the Boltzmann distribution. In liquid bilayer state, individual phosphatidylcholine lipid samples the conformational ensemble within nanosecond timescale~\cite{ferreira15}. The most accurate experimental information on conformational ensembles of lipids
in this biologically relevant phase are typically derived from NMR experiments, particularly from the
C--H bond order parameters, $S_\mathrm{CH}=\frac{3}{2} \langle \cos^2 \theta -1 \rangle$, where $\theta$ is the angle between C-H bond and membrane normal, and the average is taken over the conformational ensemble of lipids~\cite{seelig77c,davis83,Semchyschyn04}.
Because these order parameters are similar as in living cells \cite{gally81,scherer87,seelig90},
the model membranes can be used to resolve the lipid headgroup conformations in biological conditions.
According to these experiments, the glycerol backbone conformations are largely similar irrespective of the headgroup \cite{gally81}, and
the headgroup conformations are similar in the phosphatidylcholine (PC), phosphatidylethanolamine (PE), and phosphatidylglycerol (PG) lipids,
whereas the phosphatidylserine (PS) headgroup is more rigid \cite{wohlgemuth80,buldt81}. 
Notably, however, the signs of $S_\mathrm{CH}$ are not accessible from $^2$H\,NMR experiments \cite{ollila16},
and universal models to map order parameters to structural ensembles are not available \cite{pezeshkian18,akutsu20}.
Therefore, it is not clear if the lipid headgroups in a liquid lamellar bilayer exchange between a few restricted conformations, or can fluctuate freely across a wide conformational range.

@@

Our results suggest that lipid headgroup binding to proteins can be described by the {\it inverse conformational selection model}
where unbound lipids sample a wide conformational ensemble offering the large selection of conformations
that can bind to various binding sites in substrates, such as protein, drug, RNA, or virus, that remain conformationally fixed.
In this model, the binding energy is dominated by
intermolecular interactions and changes in lipid conformational entropy, while 
conformational restrictions inherent to the lipids play a minor role.
Here we have focused only on lipid headgroups while potential applications of the proposed model on interactions between hydrophobic acyl chains and proteins are left for future studies. 
Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling
and functions of lipid regulated membrane proteins in general. 
In bionanotechnology, our model facilitates, for example,
the design of phoshopolipid-specific antibodies and lipid nanoparticle carriers with minimal sides effects for mRNA vaccines.
Furthermore, our work has demonstrated the power of open access MD simulation data
to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.
